---
title: "How to create crime maps of Mexico City"
author: "Diego"
date: "October 23, 2016"
output: 
  html_document: 
    self_contained: no
---



Thanks to an information request to the SSP-CDMX
[hoyodecrimen.com](https://hoyodecrimen.com) now has crime data at the
[latitude and longitude level](https://hoyodecrimen.com/mapa#/19.36753/-99.14754/11). You can access the data systematically
using the [API](https://hoyodecrimen.com/api) or [download it](https://hoyodecrimen.com/acerca) in full. In this post I'll
provide many examples of how to manipulate the API to create maps.


## I.

A full description of all API calls is available from the
[documentation](https://hoyodecrimen.com/api). First, we'll load the R
packages we need:

```{r}
knitr::opts_chunk$set(echo = TRUE)
library("DCluster")
library("jsonlite")
library("RCurl")
library("jsonlite")
library("rgdal")
library("rgeos")
library("ggplot2")
library("ggmap")
library("RColorBrewer")
library("stringr")
library("scales")
library("geojsonio")
library("downloader")
library("spdep")
library("viridis")
library("maptools")
library("rvest")
library("dplyr")
library("stringr")
library("stringi")
library("geojsonio")
library("knitr")
```

Then, we'll download a list of all crimes available from hoyodecrimen
(since the data comes from
[FOIA request](https://hoyodecrimen.com/acerca) the specific crimes
available may vary depending on when you access the API). To do this
we can use the jsonlite package to convert the JSON data the API
endpoints return into data.frames.

```{r}
crimes <- fromJSON("https://hoyodecrimen.com/api/v1/crimes")$rows
kable(crimes, caption = 'List of crimes available from hoyodecrimen.com')
```

All requests to the API start with the address
`https://hoyodecrimen.com/api/v1/` followed by the specific path of
the data you want, sometimes with parameters (the part of the url
followed by ?) specifying the start and end dates of the data you
need.

If you wanted to download the sum of all murders commited in CDMX
since 2014 you'd look up the appropriate method in the
[documentation](https://hoyodecrimen.com/api/#get--api-v1-df-crimes-(string-crime)-series). And
then use the following code to create a chart. (Note that since the
crime we want to graph, 'Homicidio Doloso', contains a space we need
to encode the URL)

```{r}
homicidio_series <- fromJSON(str_c("https://hoyodecrimen.com",
                                   URLencode("/api/v1/df/crimes/HOMICIDIO DOLOSO/series?start_date=2014-01&end_date=2016-09")))$rows
kable(head(homicidio_series), caption = 'Homicides in CDMX since 2014')
ggplot(homicidio_series, aes(as.Date(str_c(date, "-01")), count)) +
  geom_line() +
  ggtitle("CDMX - HOMICIDIO DOLOSO") +
  xlab("date") + 
  scale_y_continuous(expand = c(0, 0), limits = c(0, max(homicidio_series$count)))

```

We can also download geojson maps of
[cuadrantes](https://hoyodecrimen.com/api/#get--api-v1-cuadrantes-geojson)
and
[sectores](https://hoyodecrimen.com/api/#get--api-v1-sectores-geojson)
which we can convert to SpatialPolygonsDataFrame to work with in R.

```{r geojson, warning=FALSE,message=FALSE}
# create a tempory to save the geojson from the API
tmp_cuadrantes = tempfile("cuads", fileext = ".json")
download("https://hoyodecrimen.com/api/v1/cuadrantes/geojson", tmp_cuadrantes)
# read the geojson into a spatial object
cuadrantes = readOGR(tmp_cuadrantes, "OGRGeoJSON", verbose = FALSE)

tmp_sectores = tempfile("secs", fileext = ".json")
download("https://hoyodecrimen.com/api/v1/sectores/geojson", tmp_sectores)
sectores = readOGR(tmp_sectores, "OGRGeoJSON", verbose = FALSE)

plot(cuadrantes, main = "cuadrantes")
plot(sectores, main = "sectores")
```

## II.

Now that we have the maps, we can download
[period data](https://hoyodecrimen.com/api/#get--api-v1-sectores-(string-sector)-crimes-(string-crime)-period)
(by default the last twelve months) for all crimes and merge
them. Then use the ggmap package to create choropleths on top of a
Google Maps image.

```{r, warning=FALSE,message=FALSE}
crime.sectors <- fromJSON("https://hoyodecrimen.com/api/v1/sectores/all/crimes/all/period")$rows
#fortify the data for ggplot2
fsectors <- fortify(sectores, region = "sector")
sector.map <- left_join(fsectors, crime.sectors, by = c("id" = "sector"))
sector.map$rate <- sector.map$count / sector.map$population * 10^5

crime.cuadrantes <- fromJSON("https://hoyodecrimen.com/api/v1/cuadrantes/all/crimes/all/period")$rows
fcuadrantes <- fortify(cuadrantes, region = "cuadrante")
cuadrante.map <- left_join(fcuadrantes, crime.cuadrantes, by = c("id" = "cuadrante"))
cuadrante.map$rate <- cuadrante.map$count / cuadrante.map$population * 10^5

draw_gmap <- function(map, crimeName, bb, pal, fill = "rate", alpha=.9) {
  ggmap(get_map(location = bb)) + 
    geom_polygon(data= subset(map, crime == crimeName), 
                 aes_string("long", "lat", group = "group", fill = fill),
                 color = "#666666", size = .1,
                 alpha = alpha) +
    coord_map() +
    ggtitle(crimeName) +
    #scale_fill_viridis(option="plasma") +
    scale_fill_continuous(low = brewer.pal(9, pal)[1],
                          high = brewer.pal(9, pal)[9],
                          space = "Lab", na.value = "grey50",
                          guide = "colourbar") +
    theme_nothing(legend = TRUE) 
}

# The bounding box for downloading the Google Maps images
bb.sector <- bbox(sectores)
draw_gmap(sector.map, "HOMICIDIO DOLOSO", bb.sector, "Reds", "rate")
draw_gmap(sector.map, "LESIONES POR ARMA DE FUEGO", bb.sector, "Greys", "rate")
draw_gmap(sector.map, "ROBO A TRANSEUNTE C.V.", bb.sector, "Greys", "rate")


draw_gmap(cuadrante.map, "ROBO DE VEHICULO AUTOMOTOR S.V.", bb.sector, "YlGn", "count")
draw_gmap(cuadrante.map, "ROBO DE VEHICULO AUTOMOTOR C.V.", bb.sector, "Greens", "count")
draw_gmap(cuadrante.map, "LESIONES POR ARMA DE FUEGO", bb.sector, "Greys", "count")
draw_gmap(cuadrante.map, "HOMICIDIO DOLOSO", bb.sector, "Reds", "count")



```

## III.

One thing that you may have noticed from the 'Homicidio Doloso' cuadrante chart is that cuadrantes with hospitals tend to have very high homicides counts (dark red in the map)

-   O-2.5.7
-   O-2.2.4
-   N-4.4.4
-   N-1.3.10
-   C-2.1.16
-   N-2.2.1
-   P-1.5.7
-   P-3.1.1

We can investigate the phenomen further by [downloading the coordinates](https://hoyodecrimen.com/api/#get--api-v1-latlong-crimes-(string-crime)-coords-(string-long)-(string-lat)-distance-(int-distance)) of all crimes within 1,000 meters of some hospitals.


```{r, warning=FALSE,message=FALSE}

#function to download a cuadrante given a latitude and longitude
get_cuadrante <- function(long, lat){
  tmp_file = tempfile(fileext = ".geojson")
  write(fromJSON(str_c("https://hoyodecrimen.com/api/v1/cuadrantes/pip/",long,"/",lat))$pip$geometry,
        file = tmp_file)
  cuad <- geojson_read(tmp_file, method = "local", what = "sp")
  cuad.f <- fortify(cuad)
}

#function to create a point map of crimes with their cuadrante
latlong_map <- function(title, lat, long, distance, crime = "HOMICIDIO DOLOSO",
                        fill = "red") {
  geocrimes <- fromJSON(str_c("https://hoyodecrimen.com/api/v1/latlong/crimes/", 
                              URLencode(crime), "/coords/",
                              long,
                              "/",
                              lat,
                              "/distance/",
                              distance,
                              "?start_date=2013-01&end_date=2016-09"))$rows
  
  cuadrante.f <- get_cuadrante(long, lat)
  
  bb <- bbox(coordinates(geocrimes[,c("long", "lat")]))
  ggmap(get_map(location = bb))  +
  geom_polygon(data = cuadrante.f, aes(long, lat, group = group), 
               color = "black", fill = "yellow", alpha = .2) + 
    geom_point(data= subset(geocrimes, crime == crime), 
               aes(long, lat),
               fill = fill,
               color = "black",
               size = 3, 
               shape = 21,
               alpha = .8) +
    ggtitle(title) +
    theme_nothing(legend = TRUE) 
}

latlong_map("Hospital Balbuena - Murder", 19.42410, -99.115520, 1000)
latlong_map("Hospitales en Lindavista - Murder", 19.482973, -99.134091, 1000)
latlong_map("Hospital General - Murder", 19.411300, -99.152405, 1000)
latlong_map("Hospital General Iztapalapa - Murder", 19.343515, -99.027382, 1000)
latlong_map("Hospital General Iztapalapa - Firearm Lesions", 
            19.343515, -99.027382, 
            1000, 
            "LESIONES POR ARMA DE FUEGO", fill = "darkgray")
```

It looks like some homicides are recorded with the latitude and
longitud of the place of death, and in the case of firearm lesions I'm
guessing it's where the crime was reported to the police.

Just for fun here are the cuadrantes with the highest counts of car robbery:

```{r, warning=FALSE,message=FALSE}
latlong_map("Mercado Xochimilo - Car Robbery wo/violence", 19.251478, -99.094207, 4900, "ROBO DE VEHICULO AUTOMOTOR S.V.", fill = "green")
latlong_map("Lindavista - Car Robbery w/violence", 19.506460, -99.122815, 3500, "ROBO DE VEHICULO AUTOMOTOR C.V.", fill = "darkgreen")
```

## IV.

Ideally we'd like to be able to map the rates of crimes at the
cuadrante delictivo level, like we did with the sectores, but the
crime rates can be statistically unstable due to the small size of the
cuadrantes and the relative small number of people at risk. Plus, as I
mentioned, homicides and firearm lesions are sometimes recorded as
having happened inside a hospital. To get a better sense of the crime
risk in each cuadrante we are going to smooth the rate using empirical
bayes smoothing.

```{r, warning=FALSE,message=FALSE}


hom <- fromJSON(str_c("https://hoyodecrimen.com",
                      URLencode("/api/v1/cuadrantes/ALL/crimes/HOMICIDIO DOLOSO/period")))$rows
hom <- subset(hom, cuadrante != "(NO ESPECIFICADO)")
# match the order with the neighborhood file
hom <- hom[match(cuadrantes$cuadrante, hom$cuadrante),]
# fill in the population of cuadrantes with zero residents
# with the mean of their neighboring cuadrantes
for(zero_cuad in hom$cuadrante[which(hom$population == 0)])
  hom$population <- mean(hom$population[cuad.nb[[which(hom$cuadrante == zero_cuad)]]])


hom$rate <- hom$count / hom$population * 10^5

smth<-empbaysmooth(hom$count, hom$population * sum(hom$count) / sum(hom$population))
hom$smooth <- smth$smthrr

cuadrantes.f <- fortify(cuadrantes, region = "cuadrante")
cuadrantes.f <- merge(cuadrantes.f, hom, by.x = "id", by.y = "cuadrante")

ggplot(cuadrantes.f, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = smooth), color = "#555555", size = .1) +
  coord_map()+
  scale_fill_gradientn(colours=rev(brewer.pal(9,"YlOrRd"))) +
  theme_nothing(legend = TRUE) +
  ggtitle("Smoothed homicide rates at the cuadrante level")
```

The cuadrantes with hospital still give the impression of being
extremely violent. If you remember the lat and long charts there were
a lot of red dots near hospitals, there is a lot of
violence near some hospitals. And looking at this
[choropleth of counts](https://hoyodecrimen.com/cuadrantes-mapa#14/19.4048/-99.1344)
makes it obvious there is a lot of violence around some hospitals (the
dark red polygons).

This makes it likely that the people who died in a hospital were injured
in a nearby location and were transported to a hospital, where they
died. A good idea would be to average the rate of violence in each
cuadrante with the nearest, say, 8 neighbors.

```{r}
cuad.nb <- knn2nb(knearneigh(coordinates(cuadrantes), k = 8), 
                  row.names = as.character(cuadrantes$cuadrante))
#cuad.nb <- poly2nb(cuadrantes, row.names = as.character(cuadrantes$cuadrante))
plot(cuad.nb, coordinates(cuadrantes), main = "Plot of cuadrante neighbors")

hom$smooth <- sapply(1:nrow(hom), function(x) {
  w <-  c(hom$population[x], hom$population[cuad.nb[[x]]])
  r <- c(hom$rate[x], hom$rate[cuad.nb[[x]]])
  return(sum(w * r)/sum(w))
})

cuadrantes.f <- fortify(cuadrantes, region = "cuadrante")
cuadrantes.f <- merge(cuadrantes.f, hom, by.x = "id", by.y = "cuadrante")

ggplot(cuadrantes.f, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = smooth), color = "#555555", size = .1) +
  coord_map()+
  scale_fill_gradientn(colours=rev(brewer.pal(9,"YlOrRd"))) +
  theme_nothing(legend = TRUE) +
  ggtitle("Smoothed homicide rates by nearest neighbors")

```

Now it looks much better and this is what the frontpage of
[hoyodecrimen](https://hoyodecrimen.com) uses to compare crime rates in your cuadrante. We can use the same
averaging method for other crimes:

```{r, warning=FALSE,message=FALSE}


smooth_cuads <- function(df, cuad.nb) {
  df <- subset(df, cuadrante != "(NO ESPECIFICADO)")
  # match the order with the neighborhood file
  df <- df[match(cuadrantes$cuadrante, df$cuadrante),]
  # fill in the population of cuadrantes with zero residents
  # with the mean of their neighboring cuadrantes
  for(zero_cuad in df$cuadrante[which(df$population == 0)])
    df$population <- mean(df$population[cuad.nb[[which(df$cuadrante == zero_cuad)]]])
  
  
  df$rate <- df$count / df$population * 10^5
  
  df$smooth <- sapply(1:nrow(df), function(x) {
    w <-  c(df$population[x], df$population[cuad.nb[[x]]])
    r <- c(df$rate[x], df$rate[cuad.nb[[x]]])
    return(sum(w * r)/sum(w))
  })
  
  cuadrantes.f <- fortify(cuadrantes, region = "cuadrante")
  cuadrantes.f <- merge(cuadrantes.f, df, by.x = "id", by.y = "cuadrante")
  cuadrantes.f
}

cuad.nb <- poly2nb(cuadrantes, row.names = as.character(cuadrantes$cuadrante))

cuad_smooth_rvsv <- smooth_cuads(fromJSON("https://hoyodecrimen.com/api/v1/cuadrantes/ALL/crimes/ROBO DE VEHICULO AUTOMOTOR S.V./period")$rows, 
             cuad.nb)

ggplot(cuad_smooth_rvsv, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = smooth), color = "#555555", size = .1) +
  coord_map()+
  scale_fill_gradientn(colours=rev(brewer.pal(9,"YlGn"))) +
  theme_nothing(legend = TRUE) +
  ggtitle("Smoothed car robbery wo/v rates by nearest neighbors")

cuad_smooth_rvcv <- smooth_cuads(fromJSON("https://hoyodecrimen.com/api/v1/cuadrantes/ALL/crimes/ROBO DE VEHICULO AUTOMOTOR C.V./period")$rows, 
             cuad.nb)

ggplot(cuad_smooth_rvcv, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = smooth), color = "#555555", size = .1) +
  coord_map() +
  scale_fill_gradientn(colours=rev(brewer.pal(9,"BuGn"))) +
  theme_nothing(legend = TRUE) +
  ggtitle("Smoothed car robbery w/v rates by nearest neighbors")

cuad_smooth_rt <- smooth_cuads(fromJSON("https://hoyodecrimen.com/api/v1/cuadrantes/ALL/crimes/ROBO A TRANSEUNTE C.V./period")$rows, 
             cuad.nb)

ggplot(cuad_smooth_rt, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = smooth), color = "#555555", size = .1) +
  coord_map() +
  scale_fill_gradientn(colours=rev(brewer.pal(9,"Blues")))+
  theme_nothing(legend = TRUE) +
  ggtitle("Smoothed street robbery rates by nearest neighbors")

```


There's still a problem in that to calculate the rates we are using as
denominator the number of people living in the cuadrante, and people
move around the city all time. Lucky for us, the Mexico City subway
website publishes the number of passangers using each metro station
which we can use to estimate crime rates by station

## V.

First, we download crime data from January to March 2016 since that
matches the ridership data available from the
[Sistema Colectivo website](http://www.metro.cdmx.gob.mx/operacion/afluencia.html).
We'll only download data for the following crimes:

* ROBO A TRANSEUNTE C.V.
* ROBO A TRANSEUNTE S.V.
* ROBO A BORDO DE TAXI C.V
* ROBO A BORDO DE MICROBUS S.V.
* ROBO A BORDO DE MICROBUS C.V.
* ROBO A BORDO DE METRO S.V.
* ROBO A BORDO DE METRO C.V.


```{r}
# Set the distance to ridiculously large number of meters to download all CDMX data
geocrimes <- fromJSON(str_c("https://hoyodecrimen.com/api/v1/latlong/crimes/", 
                            URLencode("ROBO A TRANSEUNTE C.V.,ROBO A TRANSEUNTE S.V.,ROBO A BORDO DE TAXI C.V.,ROBO A BORDO DE MICROBUS S.V.,ROBO A BORDO DE MICROBUS C.V.,ROBO A BORDO DE METRO S.V.,ROBO A BORDO DE METRO C.V."),"/coords/-99.122815/19.506460/distance/50000000000?start_date=2016-01&end_date=2016-04"))$rows
```

Then we clean the ridership data from the
[metro website](http://www.metro.cdmx.gob.mx/operacion/afluencia.html). I
had to clean up by hand the html from the website since it's so badly
formed the rvest package was unable to parse it.

```{r}
con <- file("data/afluencia.html", "rb")
metro <- read_html(con)                      


afluencia <- data.frame()
for(i in 1:4) {
  df <- metro %>% 
    html_nodes("table") %>%
    .[[i]] %>%
    html_table(fill = TRUE)
  afluencia <- rbind(afluencia,
                     data.frame(name=df[2:(nrow(df)-1),1], 
                                num=df[2:(nrow(df)-1),2],
                                line=df[1,1]),
                     data.frame(name=df[2:(nrow(df)-1),4], 
                                num=df[2:(nrow(df)-1),5],
                                line=df[1,4]),
                     data.frame(name=df[2:(nrow(df)-1),7], 
                                num=df[2:(nrow(df)-1),8],
                                line=df[1,7]))
}


nbs <- stri_escape_unicode(afluencia$name[21])
# \\u00a0 is actually the non-breaking space character
# which we have to remove from the html table
afluencia <- afluencia %>% 
  mutate(num = as.numeric(str_replace_all(num, ",", ""))) %>%
  mutate(name = str_replace_all(name, "[\r\n]" , " ")) %>%
  mutate(name = str_replace_all(name, "                       " , " ")) %>%
  mutate(name = str_replace(name, "\\u00a0", "")) %>%
  mutate(name = str_replace(name, "  ", " ")) %>%
  mutate(name = tolower(name)) %>%
  filter(name != "") %>%
  mutate(line = str_replace_all(line, "[\r\n]" , " ")) %>%
  mutate(line = str_replace_all(line, "                      " , " ")) %>%
  mutate(line = str_replace_all(line, "   " , " "))
```

Then we merge the ridership numbers with a list of geocoded stations previously
download from [here](http://www.mexicometro.org/Mexico-Metro.kmz) and converted to
a csv with the latitude and longitude of each metro station.

```{r}
stations <- read.csv("data/stations.csv") %>%
  mutate(name = tolower(name)) %>%
  rename(line = styleUrl) %>%
  mutate(line = str_replace(line, "#Line", "LÍNEA "))
df <- full_join(stations, afluencia, by = c("name", "line"))

nrow(df[is.na(df$X) | is.na(df$num),]) == 0
```

To compute the rates by station, first we are going to draw a 200m
buffer around each station point. Sometimes stations are very close to
each other and we have to be sure to merge them if they 'touch'. This can
be done by using Postgresql with the following code:

``` sql
SELECT json_build_object(
                         'type',       'Feature',
                         'id',         name,
                         'geometry',   ST_Union(the_geom::geometry),
                         'properties', json_build_object(
                                              'name', name,
                                              'num', sum(num)
                          )
                         )
FROM (SELECT ST_Buffer(the_geom_webmercator::geometry,200) as the_geom,name, line, num
      FROM geocoded_stations) as buffers
GROUP BY name
```

I've also uploaded the station points to [carto](http://www.carto.com) so you
don't have to install Postgresql and can download a geojson file of
the stations with buffer by applying some simple sql.

```{r}
url <- str_c("https://diegovalle.carto.com/api/v2/sql?format=GeoJSON&q=",
             URLencode("SELECT  ST_Transform(ST_Union(the_geom),4326) as the_geom, name, sum(num)
from (SELECT ST_Buffer(the_geom_webmercator,200) as the_geom,name, line, num
      from geocoded_stations) as buffers 
GROUP BY name"))

stations_merged <- geojson_read(url, method = "local", what = "sp")

plot(stations_merged, main = "Metro stations with a 200m buffer")


```

Now its just a matter of counting the number of crimes inside each of
the polygons taking into account the stations that were merged because
they were close to each other.

``` sql
SELECT count(crime_lat_long.the_geom), stations.name AS totale
FROM
(SELECT  ST_Transform(ST_Union(the_geom),4326) as the_geom, name, sum(num)
from (SELECT ST_Buffer(the_geom_webmercator,200) as the_geom,name, line, num
      from geocoded_stations) as buffers
GROUP BY name) as stations
LEFT JOIN crime_lat_long
ON st_contains(stations.the_geom, crime_lat_long.the_geom)
GROUP BY stations.name
```

Using carto:

```{r}
url <- str_c("https://diegovalle.carto.com/api/v2/sql?q=",
 URLencode("SELECT count(crime_lat_long.the_geom), stations.name AS name 
FROM 
(SELECT  ST_Transform(ST_Union(the_geom),4326) as the_geom, name, sum(num)
from (SELECT ST_Buffer(the_geom_webmercator,200) as the_geom,name, line, num
      from geocoded_stations) as buffers 
GROUP BY name) as stations LEFT JOIN 
crime_lat_long
ON st_contains(stations.the_geom, crime_lat_long.the_geom) 
GROUP BY stations.name"))
numcrime <-fromJSON(url)$rows


stations_merged@data <- left_join(stations_merged@data, 
                                  numcrime, 
                                  by = c("name" = "name"))
stations_merged@data$rate <- stations_merged@data$count / stations_merged@data$sum * 10^5
stations_merged <- stations_merged[!stations_merged@data$name %in% c("los reyes",
                                                                    "la paz",
                                                                    "nezahualcóyotl",
                                                                    "impulsora",
                                                                    "río de los remedios",
                                                                    "múzquiz",
                                                                    "tecnológico",
                                                                    "olímpica",
                                                                    "plaza aragón",
                                                                    "ciudad azteca",
                                                                    "cuatro caminos"), ]

writeOGR(stations_merged, 
         "html/stations.geojson", 
         driver = "GeoJSON",
         layer = "stations.geojson",
         verbose = FALSE)

map <- fortify(stations_merged, region = "name")
map <- left_join(map, stations_merged@data,by = c("id" = "name"))

bb <- bbox(stations_merged)
ggmap(get_map(location = bb)) + 
  geom_polygon(data= subset(map, rate != 0),
               color = "black",
                 aes(long, lat, group = group, fill = rate)) +
  scale_fill_continuous(low = "#ffeda0", high = "#f03b20") +
  ggtitle(title) +
  theme_nothing(legend = TRUE) +
  ggtitle("Crime by Mexico City Metro Station")

```


There's still a problem in that the crimes may be systematically biased since the
reporting rate is so low in Mexico.

You can view an interactive version [here](https://www.diegovalle.net/maps/metro/)

